---
title: "SSR_cone"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, eval=FALSE}
library(ScPoEconometrics)
launchApp("SSR_cone")
```

This app provides you with the visual intuition behind the minimization problem that Ordinary Least Squares (OLS) is looking to solve.

As you saw in `aboutApp("reg_simple")`, OLS tries to minimize the **total sum of squared errors** (SSR):

\begin{align}
SSR(b_0,b_1) &= \sum_{i = 1}^{N}(\text{observed value of }y_i - \text{predicted value of }y_i)^2 \\
&=\sum_{i = 1}^{N}(y_i - (b_0 + b_1 x_i))^2
\end{align}

Given a set of data $(x,y)$, the value of SSR is a function of $b_0,b_1$. That means, for each value of $b_0,b_1$ you choose, the output of function SSR will change. We want to find the two values for which SSR is smallest:

$$(b_0,b_1) = \arg \min_{j, s} \text{SSR}(j, s)$$

This minimization problem can be vizualized in three-dimentions by plotting the SSR as a function of $b_0$ and $b_1$. The resulting figure is called the **SSR cone** because, well, it is shaped like a cone that stands on its head. This also called a *convex* cone.


### Observe

+ Why do you think the SSR cone is shaped like this? (hint: this has to do with the "squares" part of ordinary least squares.)

+ Try and move one slider. How does your current SSR value moves along the cone?

+ Find the best values for $b_0$ and $b_1$ (the plot should turn green) and check that you are indeed at the very bottom of the SSR cone.

+ Does this minimization problem only have one unique solution? In general, does this have to be the case?











