---
title: "SSR_cone"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, eval=FALSE}
library(ScPoEconometrics)
launchApp("SSR_cone")
```

This app provides you with the visual intuition behind the minimization problem that Ordinary Least Squares (OLS) is looking to solve.

As you saw in `aboutApp("reg_simple")`, OLS tries to minimize the **total sum of squared errors** (SSR):

$$SSR = \sum_{i = 1}^{N}(\text{observed value of }y_i - \text{predicted value of }y_i)^2$$
Over two values, the intercept ($\alpha$) and the slope ($\beta$) coefficients of the regression line. In equations:

\DeclareMathOperator*{\argmin}{arg\,min}
$$(\hat\alpha, \hat\beta) = \argmin_{\hat\alpha, \hat\beta} \text{SSR}(\hat\alpha, \hat\beta)$$
or more explicitly:

$$(\hat\alpha, \hat\beta) = \argmin_{\hat\alpha, \hat\beta} \sum_{i = 1}^N(y_i - \hat\alpha - \hat\beta x_i)^2$$

This minimization problem can be vizualized in three-dimentions by plotting the SSR as a function of $\hat\alpha$ and $\hat\beta$. The resulting figure is called the **SSR cone** because, well, it is shaped like a cone.


### Observe

+ Why do you think the SSR cone is shaped like this? (hint: this has to do with the "squares" part of ordinary least squares.)

+ Try and move one slider. How does your current SSR value moves along the cone?

+ Find the best values for $\hat\alpha$ and $\hat\beta$ (the plot should turn green) and check that you are indeed at the very bottom of the SSR cone.

+ Does this minimization problem only have one unique solution? In general, does this have to be the case?











