---
title: "about `estimate`"
author: "Florian Oswald"
date: "10/30/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## about the `estimate` app

```{r,eval=FALSE}
library(ScPoEconometrics)
launchApp("estimate")
```

This app shows you what happens if we take 100 *different* samples from an normal distribution with unknown mean and variance $\mu,\sigma^2$, and we want to use those samples to **estimate** both of those *population parameters*. That is, there is an underlying normal distribution, our DGP, specified like this:

$$
\mathcal{N}(\mu,\sigma^2)
$$
We don't know the values for $\mu,\sigma^2$. We only have 100 samples that were drawn from that distribution, and from each sample we compute the associated sample mean $\bar{x}$. Each sample is of size $N$. 

## Observe

* As you increase $N$, each of the 100 samples contains more data points. $N$ points, to be precise. :-)
* You can see in the density plot, that our estimates concentrate around a certain value.
* You can see that the density more and more resembles a *normal distribution* itself!
* You can see in the table that the standard deviation of all our $\bar{x}$ gets smaller
* You can see in the table that the median of all our $\bar{x}$ gets closer to $\mu=5$.
* You can see in the table that the range of all our $\bar{x}$ gets smaller and smaller.
