---
title: "correlation_apps"
runtime: shiny_prerendered
output: html_document
---

### Continuous Correlation - Illustration

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mvtnorm)
library(ggplot2)
library(reshape2)
```

```{r, echo=FALSE}
shinyUI(fluidPage(
  sliderInput("c", "Correlation r", min = -1, max = 1, step = .05, value = 0),
  
  plotOutput("corrPlot")))
```

```{r, context="server"}

output$corrPlot <- renderPlot({
  set.seed(10)
  cor = input$c
  sig = matrix(c(1,cor,cor,1),c(2,2))
  ndat = data.frame(rmvnorm(n=300,sigma = sig))
  x = ndat$X1
  y = ndat$X2
  par(pty="s")
  plot(x ~ y, xlab="x",ylab="y",
       xlim = c(-4, 4), ylim = c(-4, 4),
       main = paste0("Correlation: ", cor),
       col = "royal blue")
})
```

### Discrete Correlation - Illustration

In this section, we throw 2 dice together many times and keep track of each of their values after each throw.

```{r, echo=FALSE}
shinyUI(fluidPage(
  sliderInput("numrep", "Number of throws", min = 1, max = 10000, step = 5, value = 10),
  
  plotOutput("outcomeTable")))
```

```{r, context="server"}

output$outcomeTable <- renderPlot({
  simulate_throw <- function(num_rep, proportions = FALSE){
  
    R <- matrix(data = 0, nrow = 6, ncol = 6) #keeps track of result
    # Arbitrary distribution
    distribution <- matrix(1/60, ncol = 6, nrow = 6)
    diag(distribution) <- 1/12
    # Simulation
    for (n in 1:num_rep){
      adder <- matrix(data = 0, nrow = 6, ncol = 6)
      adder[distribution >= matrix(runif(n = 36, min = 0, max = 1),
                                   nrow = 6, ncol = 6)] <- 1
      R <- R + adder
    }
    #Return result of the simulation
    if (proportions == FALSE){
      return(R)
    } else {return(R/num_rep)}
  }

  ggplot(melt(simulate_throw(input$numrep, T), value.name = "Frequency"), aes(Var1, Var2)) +
    geom_tile(aes(fill = Frequency), colour = "white") +
    geom_text(aes(label = paste0(round(Frequency*100, 2), "%"))) +
    scale_fill_gradient(low = "white", high = "red", guide = "colourbar") +
    xlab("First Dice") +
    ylab("Second Dice")+
    ggtitle(paste0("Outcome of ", input$numrep, " draws."), subtitle = "(in proportions)") +
    scale_y_continuous(breaks=1:6) +
    scale_x_continuous(breaks=1:6) +
    coord_fixed()

})
```



Do you think it is likely that this pattern be observed by actually throwing two fair dice many times? What do you expect this outcome to look like if you were to throw 2 fair dice yourself?

#### Here is the same app, with two *fair* dice:

```{r, echo=FALSE}
shinyUI(fluidPage(
  sliderInput("numrepf", "Number of throws", min = 1, max = 10000, step = 5, value = 10),
  
  plotOutput("outcomeTableF")))
```

```{r, context="server"}

output$outcomeTableF <- renderPlot({
  simulate_throw <- function(num_rep, proportions = FALSE){
  
    R <- matrix(data = 0, nrow = 6, ncol = 6) #keeps track of result
    # Arbitrary distribution
    distribution <- matrix(1/36, ncol = 6, nrow = 6)
    # Simulation
    for (n in 1:num_rep){
      adder <- matrix(data = 0, nrow = 6, ncol = 6)
      adder[distribution >= matrix(runif(n = 36, min = 0, max = 1),
                                   nrow = 6, ncol = 6)] <- 1
      R <- R + adder
    }
    #Return result of the simulation
    if (proportions == FALSE){
      return(R)
    } else {return(R/num_rep)}
  }

  ggplot(melt(simulate_throw(input$numrepf, T), value.name = "Frequency"), aes(Var1, Var2)) +
    geom_tile(aes(fill = Frequency), colour = "white") +
    geom_text(aes(label = paste0(round(Frequency*100, 2), "%"))) +
    scale_fill_gradient(low = "white", high = "red", guide = "colourbar") +
    xlab("First Dice") +
    ylab("Second Dice")+
    ggtitle(paste0("Outcome of ", input$numrepf, " draws."), subtitle = "(in proportions)") +
    scale_y_continuous(breaks=1:6) +
    scale_x_continuous(breaks=1:6) +
    coord_fixed()

})
```


