# Working With Data  {#sum}


In this chapter we will first learn some basic concepts that help summarizing data. Then, we will tackle a real-world task and read, clean, and summarize data from the web.

## Summary Statistics

`R` has built in functions for a large number of summary statistics. For numeric variables, we can summarize data by looking at their center and spread, for example.

```{r}
# for the mpg dataset, we load:
library(ggplot2)
```

### Central Tendency {-}

Suppose we want to know the *mean* and *median* of all the values stored in the `data.frame` column `mpg$cty`:

| Measure | `R`               | Result              |
|:---------:|:-------------------:|:---------------------:|
| Mean    | `mean(mpg$cty)`   | `r mean(mpg$cty)`   |
| Median  | `median(mpg$cty)` | `r median(mpg$cty)` |

### Spread {-}

How do the values in that column *vary*? How far *spread out* are they?

| Measure            | `R`              | Result             |
|:---------:|:-------------------:|:---------------------:|
| Variance           | `var(mpg$cty)`   | `r var(mpg$cty)`   |
| Standard Deviation | `sd(mpg$cty)`    | `r sd(mpg$cty)`    |
| IQR                | `IQR(mpg$cty)`   | `r IQR(mpg$cty)`   |
| Minimum            | `min(mpg$cty)`   | `r min(mpg$cty)`   |
| Maximum            | `max(mpg$cty)`   | `r max(mpg$cty)`   |
| Range              | `range(mpg$cty)` | `r range(mpg$cty)` |

### Categorical {-}

For categorical variables, counts and percentages can be used for summary.

```{r}
table(mpg$drv)
table(mpg$drv) / nrow(mpg)
```

## Plotting

Now that we have some data to work with, and we have learned about the data at the most basic level, our next tasks will be to visualize it. Often, a proper visualization can illuminate features of the data that can inform further analysis.

We will look at four methods of visualizing data by using the basic `plot` facilities built-in with `R`:

- Histograms
- Barplots
- Boxplots
- Scatterplots

### Histograms

When visualizing a single numerical variable, a **histogram** is useful. It summarizes the *distribution* of values in a vector. In `R` you create one using the `hist()` function:

```{r}
hist(mpg$cty)
```

The histogram function has a number of parameters which can be changed to make our plot look much nicer. Use the `?` operator to read the documentation for the `hist()` to see a full list of these parameters.

```{r}
hist(mpg$cty,
     xlab   = "Miles Per Gallon (City)",
     main   = "Histogram of MPG (City)", # main title
     breaks = 12,   # how many breaks?
     col    = "red",
     border = "blue")
```

Importantly, you should always be sure to label your axes and give the plot a title. The argument `breaks` is specific to `hist()`. Entering an integer will give a suggestion to `R` for how many bars to use for the histogram. By default `R` will attempt to intelligently guess a good number of `breaks`, but as we can see here, it is sometimes useful to modify this yourself.

### Barplots

Somewhat similar to a histogram, a barplot can provide a visual summary of a categorical variable, or a numeric variable with a finite number of values, like a ranking from 1 to 10.

```{r}
barplot(table(mpg$drv))
```

```{r}
barplot(table(mpg$drv),
        xlab   = "Drivetrain (f = FWD, r = RWD, 4 = 4WD)",
        ylab   = "Frequency",
        main   = "Drivetrains",
        col    = "dodgerblue",
        border = "darkorange")
```

### Boxplots

To visualize the relationship between a numerical and categorical variable, once could use a **boxplot**. In the `mpg` dataset, the `drv` variable takes a small, finite number of values. A car can only be front wheel drive, 4 wheel drive, or rear wheel drive.

```{r}
unique(mpg$drv)
```

First note that we can use a single boxplot as an alternative to a histogram for visualizing a single numerical variable. To do so in `R`, we use the `boxplot()` function. The box shows the *interquartile range*, the solid line in the middle is the value of the median, the wiskers show 1.5 times the interquartile range, and the dots are outliers.

```{r}
boxplot(mpg$hwy)
```

However, more often we will use boxplots to compare a numerical variable for different values of a categorical variable.

```{r}
boxplot(hwy ~ drv, data = mpg)
```

Here used the `boxplot()` command to create side-by-side boxplots. However, since we are now dealing with two variables, the syntax has changed. The `R` syntax `hwy ~ drv, data = mpg` reads "Plot the `hwy` variable against the `drv` variable using the dataset `mpg`." We see the use of a `~` (which specifies a formula) and also a `data = ` argument. This will be a syntax that is common to many functions we will use in this course. 

```{r}
boxplot(hwy ~ drv, data = mpg,
     xlab   = "Drivetrain (f = FWD, r = RWD, 4 = 4WD)",
     ylab   = "Miles Per Gallon (Highway)",
     main   = "MPG (Highway) vs Drivetrain",
     pch    = 20,
     cex    = 2,
     col    = "darkorange",
     border = "dodgerblue")
```

Again, `boxplot()` has a number of additional arguments which have the ability to make our plot more visually appealing.

### Scatterplots

Lastly, to visualize the relationship between two numeric variables we will use a **scatterplot**. This can be done with the `plot()` function and the `~` syntax we just used with a boxplot. (The function `plot()` can also be used more generally; see the documentation for details.)

```{r}
plot(hwy ~ displ, data = mpg)
```

```{r}
plot(hwy ~ displ, data = mpg,
     xlab = "Engine Displacement (in Liters)",
     ylab = "Miles Per Gallon (Highway)",
     main = "MPG (Highway) vs Engine Displacement",
     pch  = 20,
     cex  = 2,
     col  = "dodgerblue")
```

### `ggplot` {#ggplot}

All of the above plots could also have been generated using the `ggplot` function from the already loaded `ggplot2` package. Which function you use is up to you, but sometimes a plot is easier to build in base R (like in the `boxplot` example maybe), sometimes the other way around.

```{r}
ggplot(data = mpg,mapping = aes(x=displ,y=hwy)) + geom_point()
```

`ggplot` is impossible to describe in brief terms, so please look at [the package's website](http://ggplot2.tidyverse.org) which provides excellent guidance. We will from time to time use ggplot in this book, so you could familiarize yourself with it. Let's quickly demonstrate how one could further customize that first plot:

```{r}
ggplot(data = mpg, mapping = aes(x=displ,y=hwy)) +   # ggplot() makes base plot
  geom_point(color="blue",size=2) +     # how to show x and y?
  scale_y_continuous(name="Miles Per Gallon (Highway)") +  # name of y axis
  scale_x_continuous(name="Engine Displacement (in Liters)") + # x axis
  theme_bw() +    # change the background
  ggtitle("MPG (Highway) vs Engine Displacement")   # add a title
```

If you want to see `ggplot` in action, you could start with [this](http://jcyhong.github.io/ggplot_demo.html) and then look at that [very nice tutorial](https://tutorials.iq.harvard.edu/R/Rgraphics/Rgraphics.html)? It's fun!

## Summarizing Two Variables {#summarize-two}

We often are interested in how two variables are related to each other. The core concepts here are *covariance* and *correlation*. Let's generate some data on `x` and `y` and plot them against each other:

```{r x-y-corr,echo=FALSE,message=FALSE,warning=FALSE,fig.cap='How are $x$ and $y$ related?',fig.align='center'}
library(mvtnorm)
set.seed(10)
cor = 0.9
sig = matrix(c(1,cor,cor,1),c(2,2))
ndat = data.frame(rmvnorm(n=300,sigma = sig))
x = ndat$X1
y = ndat$X2
par(pty="s")
plot(x ~ y, xlab="x",ylab="y")
```

Taking as example the data in this plot, the concepts *covariance* and *correlation* relate to the following type of question:

```{block, type="note"}
Given we observe value of something like $x=2$, say, can we expect a high or a low value of $y$, on average? Something like $y=2$ or rather something like $y=-2$?
```
<br>
The answer to this type of question can be addressed by computing the covariance of both variables:

```{r}
cov(x,y)  
```

Here, this gives a positive number, `r round(cov(x,y),2)`, indicating that as one variable lies above it's average, the other one does as well. In other words, it indicates a **positive relationship**. What is less clear, however, how to interpret the magnitude of `r round(cov(x,y),2)`. Is that a *strong* or a *weak* positive association?

In fact, we cannot tell. This is because the covariance is measured in the same units as the data, and those units often differ between both variables. There is a better measure available to us though, the **correlation**, which is obtained by *standardizing* each variable. By *standardizing* a variable $x$ one means to divide $x$ by its standard deviation $\sigma_x$:

$$
z = \frac{x}{\sigma_x}
$$

The *correlation coefficient* between $x$ and $y$, commonly denoted $r_{x,y}$, is then defined as

$$
r_{x,y} = \frac{cov(x,y)}{\sigma_x \sigma_y},
$$

and we get rid of the units problem. In `R`, you can call directly

```{r}
cor(x,y)
```

Now this is better. Given that the correlation has to lie in $[-1,1]$, a value of `r round(cor(x,y),2)` is indicative of a rather strong positive relationship for the data in figure \@ref(fig:x-y-corr)

Note that $x,y$ being drawn from a *continuous distribution* (they are joint normally distributed) had no implication for covariance and correlation: We can compute those measures also for discrete random variables (like the throws of two dice, as you will see in one of our tutorials).

### Visually estimating $\sigma$

Sometimes it is useful to estimate the standard deviation of some data *without* the help of a computer (for example during an exam ;-) ). If $x$ is approximately normally distributed, 95% of its observations will lie within a range of $\bar{x}\pm$ two standard deviations of $x$. That is to say, *four* standard deviations of $x$ cover 95% of its observations. Hence, a simple way to estimate the standard deviation for a variable is to look at the range of $x$, and simply divide that number by four. 
 
```{r vis,fig.cap='visual estimation on $\\sigma$. The x-axis labels min and max as well as mean of $x$.',echo=FALSE}
sdd = 3
md = 3
dta = rnorm(50,mean=md,sd=sdd)
plot(dta,rep(1,50),pch=3,yaxt="n",ylab="",xlab="x",xaxt="n")
axis(1,at=round(c(min(dta),md,max(dta)),2))
```

This is illustrated in figure \@ref(fig:vis). Here we see that `range(x)/4` gives `r round(diff(range(dta))/4,2)` which compares favourably to the actual standard deviation `r sdd`.


## The `tidyverse`

[Hadley Wickham](http://hadley.nz) is the author of R packages `ggplot2` and also of `dplyr` (and also a myriad of others). With `ggplot2` he introduced what is called the *grammar of graphics* (hence, `gg`) to `R`. Grammar in the sense that there are **nouns** and **verbs** and a **syntax**, i.e. rules of how nouns and verbs are to be put together to construct an understandable sentence. He has extended the *grammar* idea into various other packages. The `tidyverse` package is a collection of those packages. 

`tidy` data is data where:

* Each variable is a column
* Each observation is a row
* Each value is a cell

Fair enough, you might say, that is a regular spreadsheet. And you are right! However, data comes to us *not* tidy most of the times, and we first need to clean, or `tidy`, it up. Once it's in `tidy` format, we can use the tools in the `tidyverse` with great efficiency to analyse the data and stop worrying about which tool to use.

### Reading `.csv` data in the *tidy* way

We could have used the `read_csv()` function from the `readr` package to read our example dataset from the previous chapter. The `readr` function `read_csv()` has a number of advantages over the built-in `read.csv`. For example, it is much faster reading larger data. [It also uses the `tibble` package to read the data as a tibble.](https://cran.r-project.org/web/packages/tibble/vignettes/tibble.html) **A `tibble` is simply a data frame that prints with sanity.** Notice in the output below that we are given additional information such as dimension and variable type.

```{r, message = FALSE, warning = FALSE}
library(readr)  # you need `install.packages("readr")` once!
path = system.file(package="ScPoEconometrics","datasets","example-data.csv")
example_data_from_disk = read_csv(path)
```


### Tidy `data.frames` are `tibbles`

Let's grab some data from the `ggplot2` package:

```{r}
data(mpg,package = "ggplot2")  # load dataset `mpg` from `ggplot2` package
head(mpg, n = 10)
```

The function `head()` will display the first `n` observations of the data frame, as we have seen. The `head()` function was more useful before tibbles. Notice that `mpg` is a tibble already, so the output from `head()` indicates there are only `10` observations. Note that this applies to `head(mpg, n = 10)` and not `mpg` itself. Also note that tibbles print a limited number of rows and columns by default. The last line of the printed output indicates with rows and columns were omitted.

```{r}
mpg
```

Let's look at `str` as well to get familiar with the content of the data:

```{r}
str(mpg)
```

In this dataset an observation is for a particular model-year of a car, and the variables describe attributes of the car, for example its highway fuel efficiency.

To understand more about the data set, we use the `?` operator to pull up the documentation for the data.

```{r, eval = FALSE}
?mpg
```

Working with tibbles is mostly the same as working with plain data.frames:

```{r}
names(mpg)
mpg$year
mpg$hwy
```

Subsetting is also similar to dataframe. Here, we find fuel efficient vehicles earning over 35 miles per gallon and only display `manufacturer`, `model` and `year`.

```{r}
# mpg[row condition, col condition]
mpg[mpg$hwy > 35, c("manufacturer", "model", "year")]
```

An alternative would be to use the `subset()` function, which has a much more readable syntax.

```{r, eval = FALSE}
subset(mpg, subset = hwy > 35, select = c("manufacturer", "model", "year"))
```

Lastly, and most *tidy*, we could use the `filter` and `select` functions from the `dplyr` package which introduces the *pipe operator* `f(x) %>% g(z)` from the `magrittr` package. This operator takes the output of the first command, for example `y = f(x)`, and passes it *as the first argument* to the next function, i.e. we'd obtain `g(y,z)` here.^[A *pipe* is a concept from the Unix world, where it means to take the output of some command, and pass it on to another command. This way, one can construct a *pipeline* of commands. For additional info on the pipe operator in R, you might be interested [in this tutorial](https://www.datacamp.com/community/tutorials/pipe-r-tutorial).]

```{r, eval = TRUE,message=FALSE,warning=FALSE}
library(dplyr)
mpg %>% 
  filter(hwy > 35) %>% 
  select(manufacturer, model, year)
```

Note that the above syntax is equivalent to the following pipe-free command (which is much harder to read!):

```{r, eval = TRUE,message=FALSE,warning=FALSE}
library(dplyr)
select(filter(mpg, hwy > 35), manufacturer, model, year)
```

All three approaches produce the same results. Which you use will be largely based on a given situation as well as your preference.

#### Task 1

1. Make sure to have the `mpg` dataset loaded by typing `data(mpg)` (and `library(ggplot2)` if you haven't!). Use the `table` function to find out how many cars were built by *mercury*? 
1. What is the average year the audi's were built in this dataset? Use the function `mean` on the subset of column `year` that corresponds to `audi`. (Be careful: subsetting a `tibble` returns a `tibble` (and not a vector)!. so get the `year` column after you have subset the `tibble`.) 
1. Use the `dplyr` piping syntax from above first with `group_by` and then with `summarise(newvar=your_expression)` to find the mean `year` by all manufacturers (i.e. same as previous task, but for all manufacturers. don't write a loop!).



### Tidy Example: Importing Non-Tidy Excel Data

The data we will look at is from [Eurostat](http://ec.europa.eu/eurostat/data/database) on demography and migration. You should download the data yourself (click on previous link, then drill down to *database by themes > Population and social conditions > Demograph and migration > Population change - Demographic balance and crude rates at national level (demo_gind)*). 

Once downloaded, we can read the data with the function `read_excel` from the package [`readxl`](http://readxl.tidyverse.org), again part of the `tidyverse` suite.

It's important to know how the data is organized in the spreadsheet. Open the file with Excel to see:

* There is a heading which we don't need.
* There are 5 rows with info that we don't need.
* There is one table per variable (total population, males, females, etc)
* Each table has one row for each country, and one column for each year.
* As such, this data is **not tidy**.

Now we will read the first chunk of data, from the first table: *total population*:

```{r,message=FALSE,warning=FALSE}
library(readxl)  # load the library
# Notice that if you installed the R package of this book,
# you have the .xls data file already at 
# `system.file(package="ScPoEconometrics",
#                        "datasets","demo_gind.xls")`
# otherwise:
# * download the file to your computer
# * change the argument `path` to where you downloaded it
# you may want to change your working directory with `setwd("your/directory")
# or in RStudio by clicking Session > Set Working Directory

# total population in raw format
tot_pop_raw = read_excel(
                path = system.file(package="ScPoEconometrics",
                                    "datasets","demo_gind.xls"), 
                sheet="Data", # which sheet
                range="A9:K68")  # which excel cell range to read
names(tot_pop_raw)[1] <- "Country"   # lets rename the first column
tot_pop_raw
```

This shows a `tibble`, which we encountered just above. The column names are `Country,2008,2009,...`, and the rows are numbered `1,2,3,...`. Notice, in particular, that *all* columns seem to be of type `<chr>`, i.e. characters - a string, not a number! We'll have to fix that, as this is clearly numeric data.

#### `tidyr`

In the previous `tibble`, each year is a column name (like `2008`) instead of all years being collected in one column `year`. We really would like to have several rows for each Country, one row per year. We want to `gather()` all years into a new column to tidy this up - and here is how:

1. specify which columns are to be gathered: in our case, all years (note that `paste(2008:2017)` produces a vector like `["2008", "2009", "2010",...]`)
1. say what those columns should be gathered into, i.e. what is the *key* for those values: we'll call it `year`.
1. Finally, what is the name of the new resulting column, containing the *value* from each cell: let's call it `counts`.

```{r gather,warning=FALSE}
library(tidyr)   # for the gather function
tot_pop = gather(tot_pop_raw, paste(2008:2017),key="year", value = "counts")
tot_pop
```

That's better! However, `counts` is still `chr`! Let's convert it to a number:

```{r convert}
tot_pop$counts = as.integer(tot_pop$counts)
tot_pop
```

Now you can see that column `counts` is indeed `int`, i.e. an integer number, and we are fine. The `Warning: NAs introduced by coercion` means that `R` converted some values to `NA`, because it couldn't convert them into `numeric`. More below!

#### `dplyr`

>The [transform](http://r4ds.had.co.nz/transform.html) chapter of Hadley Wickham's book is a great place to read up more on using `dplyr`.

With `dplyr` you can do the following operations on `data.frame`s and `tibble`s:

* Choose observations based on a certain value (i.e. subset): `filter()`
* Reorder rows: `arrange()`
* Select variables by name: `select()`
* Create new variables out of existing ones: `mutate()`
* Summarise variables: `summarise()`

All of those verbs can be used with `group_by()`, where we apply the respective operation on a *group* of the dataframe/tibble. For example, on our `tot_pop` tibble we will now

* filter
* mutate
* and plot the resulting values

Let's get a plot of the populations of France, the UK and Italy over time, in terms of millions of people. We will make use of the `piping` syntax of `dplyr` which we introduced just above.

```{r gather-plot,warning=FALSE,message=FALSE}
library(dplyr)  # for %>%, filter, mutate, ...
# 1. take the data.frame `tot_pop`
tot_pop %>%
  # 2. pipe it into the filter function
  # filter on Country being one of "France","United Kingdom" or "Italy"
  filter(Country %in% c("France","United Kingdom","Italy")) %>%
  # 3. pipe the result into the mutate function
  # create a new column called millions
  mutate(millions = counts / 1e6) %>%
  # 4. pipe the result into ggplot to make a plot
  ggplot(mapping = aes(x=year,y=millions,color=Country,group=Country)) + geom_line(size=1)
```

#### Arrange a `tibble` {-} 

* What are the top/bottom 5 most populated areas?

```{r,message=FALSE}
top5 = tot_pop %>%
  arrange(desc(counts)) %>%  # arrange in descending order of col `counts`
  top_n(5)

bottom5 = tot_pop %>%
  arrange(desc(counts)) %>%
  top_n(-5)
# let's see top 5
top5
# and bottom 5
bottom5
```

Now this is not exactly what we wanted. It's always the same country in both top and bottom, because there are multiple years per country. Let's compute average population over the last 5 years and rank according to that:

```{r,message=FALSE}
topbottom = tot_pop %>%
  group_by(Country) %>%
  filter(year > 2012) %>%
  summarise(mean_count = mean(counts)) %>%
  arrange(desc(mean_count))

top5 = topbottom %>% top_n(5)
bottom5 = topbottom %>% top_n(-5)
top5
bottom5
```
That's better! 

#### Look for `NA`s in a `tibble` {-} 

Sometimes data is *missing*, and `R` represents it with the special value `NA` (not available). It is good to know where in our dataset we are going to encounter any missing values, so the task here is: let's produce a table that has three columns:

1. the names of countries with missing data
2. how many years of data are missing for each of those
3. and the actual years that are missing

```{r}
missings = tot_pop %>%
  filter(is.na(counts)) %>% # is.na(x) returns TRUE if x is NA
  group_by(Country) %>%
  summarise(n_missing = n(),years = paste(year,collapse = ", "))
knitr:::kable(missings)  # knitr:::kable makes a nice table
```


#### Males and Females {-} 

Let's look at the numbers by male and female population. They are in the same xls file, but at different cell ranges. Also, I just realised that the special character `:` indicates *missing* data. We can feed that to `read_excel` and that will spare us the need to convert data types afterwards. Let's see:

```{r females}
females_raw = read_excel(
                path = system.file(package="ScPoEconometrics",
                                    "datasets","demo_gind.xls"), 
                sheet="Data", # which sheet
                range="A141:K200",  # which excel cell range to read
                na=":" )   # missing data indicator
names(females_raw)[1] <- "Country"   # lets rename the first column
females_raw
```

You can see that `R` now correctly read the numbers as such, after we told it that the `:` character has the special *missing* meaning: before, it *coerced* the entire `2008` column (for example) to be of type `chr` after it hit the first `:`. We had to manually convert the column back to `numeric`, in the process automatically coercing the `:`s into `NA`. Now we addressed that issue directly. Let's also get the male data in the same way:

```{r males}
males_raw = read_excel(
                path = system.file(package="ScPoEconometrics",
                                    "datasets","demo_gind.xls"), 
                sheet="Data", # which sheet
                range="A75:K134",  # which excel cell range to read
                na=":" )   # missing data indicator
names(males_raw)[1] <- "Country"   # lets rename the first column
```

Next step was to `tidy` up this data, just as before:

```{r tidymales}
females = gather(females_raw, paste(2008:2017),key="year", value = "counts")
males = gather(males_raw, paste(2008:2017),key="year", value = "counts")
```

Let's try to tweak our above plot to show the same data in two separate panels: one for males and one for females. This is easiest to do with `ggplot` if we have all the data in one single `data.frame` (or `tibble`), and marked with a *group identifier*. Let's first add this to both datasets, and then let's just combine both into one:

```{r}
females$sex = "female"
males$sex = "male"
sexes = rbind(males,females)   # "row bind" 2 data.frames
sexes
```

Now that we have all the data nice and `tidy` in a `data.frame`, this is a very small change to our previous plotting code:

```{r psexes}
sexes %>%
  filter(Country %in% c("France","United Kingdom","Italy")) %>%
  mutate(millions = counts / 1e6) %>%
  ggplot(mapping = aes(x=as.Date(year,format="%Y"),  # convert to `Date`
                       y=millions,colour=Country,group=Country)) + 
      geom_line() +
  scale_x_date(name = "year") + # rename x axis
  facet_wrap(~sex)   # make two panels, splitting by groups `sex`
```

#### Always Compare to Germany :-) {-}

How do our three countries compare with respect to the biggest country in the EU in terms of population? What *fraction* of Germany does the French population make in any given year, for example?

```{r}
# remember that the pipe operator %>% takes the 
# result of the previous operation and passes it
# as the *first* argument to the next function call
merge_GER <- tot_pop %>%
  # 1. subset to countries of interest
  filter(
    Country %in% 
      c("France",
        "United Kingdom",
        "Italy")
    ) %>%
  # 2. group data by year
  group_by(year) %>%
  # 3. add GER's count as new column *by year*
  left_join(
    # Germany only
    filter(tot_pop,
           Country %in% "Germany including former GDR"),
    # join back in `by year`
    by="year")
merge_GER
```
 
Here you see that the merge (or join) operation labelled `col.x` and `col.y` if
 both datasets contained a column called `col`. Now let's continue to compute what proportion of german population each country amounts to:


```{r}
names(merge_GER)[1] <- "Country"
merge_GER %>%
  mutate(prop_GER = 100 * counts.x / counts.y) %>%
  # 5. plot
  ggplot(mapping = 
           aes(x = year,
               y = prop_GER,
               color = Country,
               group = Country)) + 
  geom_line(size=1) + 
  scale_y_continuous("percent of German population") + 
  theme_bw()  # new theme for a change?
```




