# Advanced `R` {#R-advanced}

This chapter continues with some advanced usage examples from chapter \@ref(R-intro)

## More Vectorization

```{r}
x = c(1, 3, 5, 7, 8, 9)
y = 1:100
```

```{r}
x + 2
x + rep(2, 6)
```

```{r}
x > 3
x > rep(3, 6)
```

```{r}
x + y
length(x)
length(y)
length(y) / length(x)
(x + y) - y
```

```{r}
y = 1:60
x + y
length(y) / length(x)
```

```{r}
rep(x, 10) + y
```

```{r}
all(x + y == rep(x, 10) + y)
identical(x + y, rep(x, 10) + y)
```

```{r}
# ?any
# ?all.equal
```



## Calculations with Vectors and Matrices

Certain operations in `R`, for example `%*%` have different behavior on vectors and matrices. To illustrate this, we will first create two vectors.

```{r}
a_vec = c(1, 2, 3)
b_vec = c(2, 2, 2)
```

Note that these are indeed vectors. They are not matrices.

```{r}
c(is.vector(a_vec), is.vector(b_vec))
c(is.matrix(a_vec), is.matrix(b_vec))
```

When this is the case, the `%*%` operator is used to calculate the **dot product**, also know as the **inner product** of the two vectors.

The dot product of vectors $\boldsymbol{a} = \lbrack a_1, a_2, \cdots a_n \rbrack$ and $\boldsymbol{b} = \lbrack b_1, b_2, \cdots b_n \rbrack$ is defined to be

\[
\boldsymbol{a} \cdot \boldsymbol{b} = \sum_{i = 1}^{n} a_i b_i = a_1 b_1 + a_2 b_2 + \cdots a_n b_n.
\]

```{r}
a_vec %*% b_vec # inner product
a_vec %o% b_vec # outer product
```

The `%o%` operator is used to calculate the **outer product** of the two vectors.

When vectors are coerced to become matrices, they are column vectors. So a vector of length $n$ becomes an $n \times 1$ matrix after coercion.

```{r}
as.matrix(a_vec)
```

If we use the `%*%` operator on matrices, `%*%` again performs the expected matrix multiplication. So you might expect the following to produce an error, because the dimensions are incorrect.

```{r}
as.matrix(a_vec) %*% b_vec
```

At face value this is a $3 \times 1$ matrix, multiplied by a $3 \times 1$ matrix. However, when `b_vec` is automatically coerced to be a matrix, `R` decided to make it a "row vector", a $1 \times 3$ matrix, so that the multiplication has conformable dimensions.

If we had coerced both, then `R` would produce an error.

```{r, eval = FALSE}
as.matrix(a_vec) %*% as.matrix(b_vec)
```

Another way to calculate a *dot product* is with the `crossprod()` function. Given two vectors, the `crossprod()` function calculates their dot product. The function has a rather misleading name.

```{r}
crossprod(a_vec, b_vec)  # inner product
tcrossprod(a_vec, b_vec)  # outer product
```

These functions could be very useful later. When used with matrices $X$ and $Y$ as arguments, it calculates

\[
X^\top Y.
\]

When dealing with linear models, the calculation

\[
X^\top X
\]

is used repeatedly.

```{r}
C_mat = matrix(c(1, 2, 3, 4, 5, 6), 2, 3)
D_mat = matrix(c(2, 2, 2, 2, 2, 2), 2, 3)
```

This is useful both as a shortcut for a frequent calculation and as a more efficient implementation than using `t()` and `%*%`.

```{r}
crossprod(C_mat, D_mat)
t(C_mat) %*% D_mat
all.equal(crossprod(C_mat, D_mat), t(C_mat) %*% D_mat)
```

```{r}
crossprod(C_mat, C_mat)
t(C_mat) %*% C_mat
all.equal(crossprod(C_mat, C_mat), t(C_mat) %*% C_mat)
```


## Matrices


```{r}
Z = matrix(c(9, 2, -3, 2, 4, -2, -3, -2, 16), 3, byrow = TRUE)
Z
solve(Z)
```

To verify that `solve(Z)` returns the inverse, we multiply it by `Z`. We would expect this to return the identity matrix, however we see that this is not the case due to some computational issues. However, `R` also has the `all.equal()` function which checks for equality, with some small tolerance which accounts for some computational issues. The `identical()` function is used to check for exact equality.

```{r}
solve(Z) %*% Z
diag(3)
all.equal(solve(Z) %*% Z, diag(3))
```

`R` has a number of matrix specific functions for obtaining dimension and summary information.

```{r}
X = matrix(1:6, 2, 3)
X
dim(X)
rowSums(X)
colSums(X)
rowMeans(X)
colMeans(X)
```

The `diag()` function can be used in a number of ways. We can extract the diagonal of a matrix.

```{r}
diag(Z)
```

Or create a matrix with specified elements on the diagonal. (And `0` on the off-diagonals.)

```{r}
diag(1:5)
```

Or, lastly, create a square matrix of a certain dimension with `1` for every element of the diagonal and `0` for the off-diagonals.

```{r}
diag(5)
```
