---
title: "datasaurus"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, eval=FALSE}
library(ScPoEconometrics)
launchApp("datasaurus")
```

This app let's you visualize the line of best fit for 13 different x-y datasets, as well as their correlation.

### Observe

These datasets were produced by [Alberto Cairo](http://www.thefunctionalart.com/2016/08/download-datasaurus-never-trust-summary.html) in order to illustrate why you should never trust summary statistics alone and **always plot your data**. Believe it or not, all 13 datasets have:

+ Exactly the same x and y means, `mean(x)` = $54.26$ and `mean(y)` = $47.83$

+ Exactly the same x and y standard deviations, `sd(x)` = $16.76$ and `sd(y)` = $26.96$

+ Exactly the same correlation coefficient, `cor(x, y)` = $-0.06$. 

Therefore it follows that they also must have the same covariance. Can you see why?


### Explore the datasaurus yourself

You don't have to believe everything we tell you! If you wish to check the above statements for yourself, we've downloaded the data on your computer along with the class package. To get it, simply run:

```{r, eval = FALSE}
df = datasauRus::datasaurus_dozen #load the data in the `df` variable
df
```

Have fun!
